---
title: "Longitudinal Data Analysis General Framework"
author: "Yue Harriet Huang"
date: "March 25, 2016"
output: html_document
---
install.packages("reshape2")
install.packages("data.table")
install.packages("lattice")
install.packages("ggplot2")

## Design:

### * Repeated Measures: response measured repeatedly on a set of units (individuals)
### * Balanced Dataset: each individual measured at the same set of ages
### * Unbalanced dataset: each individual measured at different set of ages

## Research questions we can answer:

### * How does change of treatment (predictors) affect growth over time within the same individual

### * How does change of treatment (predictors) affect growth between individuals (Control Group versus Test Group)

```{r, echo=FALSE}

#=== import data ===#

path = "C:/Users/Yue/Documents/meetup_doc/oil_price_unemployment/TorontoMachineLearningBookClub/data"

setwd(path)

unemp = read.csv('Unemployment_Rate_alberta.csv')

#head(unemp)

#=== data only 15 years and over, Both sexes, after 2000 ===#

pattern1 = '.over'
ind1 = grep(pattern1, unemp$AgeGroup, perl = T)
subset = unemp[ind1,]

pattern2 = 'Both.'
ind2 = grep(pattern2, subset$Sex, perl=T)
subset = subset[ind2,]

subset = subset[grep("20", subset$When),]

subset$When = as.Date(as.character(subset$When), format = "%m/%d/%Y")

#=== reshape wide to long ===#

require(reshape2)
require(data.table)

subset = melt(subset, id.vars = c("When", "AgeGroup", "Sex"))

setnames(subset, "variable", "Province")
setnames(subset, "value", "Unemp")


```

```{r, echo=FALSE, fig.width=10, fig.height=10}

#=== growth curve plot ===#

require(lattice)

xyplot(Unemp ~ When | factor(Province), data=subset, as.table=T)

interaction.plot(subset$When, factor(subset$Province), subset$Unemp, col=(1:length(unique(subset$Province))), xlab="When", ylab="Unemployment Rate", main="Unemployment in Different Provinces")

par(mfrow=c(1,2))
plot(subset[subset$Province=="Alberta", 'Unemp'], main="Alberta", ylab = "Unemployment Rate", xlab="From 2000 to 2016 by Month")

plot(subset[subset$Province=="Ontario", 'Unemp'], main="Ontario", ylab = "Unemployment Rate", xlab="From 2000 to 2016 by Month")

dev.off()
```

## Observe from plots:

### 1. Variability within individuals: usually correlated
### 2. Variability between individuals (average quantity varies across individuals)


## 1. Model Variability within individual over time: 1st level model

### (1) Firstly try a linear model on each of the growth trajectory:
#### Recall Assumptions of OLS linear model:

    * independent between observations
    
#### which apparently does not make sense: 
    
    * clear nonlinearity in each of the growth curve
    
    * observations within an individual are correlated
    
    * but we can see that each curve has different intercept and slope

```{r, echo=FALSE}

xyplot(Unemp ~ When | factor(Province), data=subset, 
  panel = function(x, y){
    panel.xyplot(x, y)
    panel.lmline(x, y)
  }, as.table=T)

          

```

## Choose Fixed Effects Model or Random Effects Model

### Fixed Effects Model: 

* Fixed model cannot generalize to new individuals in the population, only to 'new' observations from the same individuals. 
In our case, this is appropriate, because Canada has a fixed number of provinces, we do not need to generalize to newer individuals (provinces) outside of our sample, otherwise we will use a Random effects model to generalized to the population.



#### 1. Level 1: modeling growth within individual: 
## $$y_{it} = \beta_{i0} + \beta_{i1}X_{it} + \epsilon_{it}$$
#### 2. Level 2: modeling difference between individuals: allowing different $\beta_{i0}$ (intercept / level) and $\beta_{i1}$ (slope / rate of change) for different groups of individuals

* From the sphaghetti plot above, we clearly see that whether it is a maritime (Newfoundland, PEI, NB, NS) - highest, or an eastern (Quebec, Ontario) - 2nd highest or a prairie (Saskatchenwan, Alberta, Manitoba) -- lowest or western province (BC) - 2nd lowest affects the level (intercept) of the plot, however, not so much of the slope (rate of change), so we model the slope as :

## $$\beta_{i0} = \phi_{00} + \phi_{01}Location_{i} + \zeta_{0i}$$

```{r, echo=FALSE}

#=== Code Location ===#

subset[subset$Province == "Alberta", "Location"] = "Prairie"
subset[subset$Province == "BritishColumbia", "Location"] = "BC"
subset[subset$Province == "Manitoba", "Location"] = "Prairie"
subset[subset$Province == "NewBrunswick", "Location"] = "Maritime"
subset[subset$Province == "NewfoundlandAndLabrador", "Location"] = "Maritime"
subset[subset$Province == "NovaScotia", "Location"] = "Maritime"
subset[subset$Province == "Ontario", "Location"] = "East"
subset[subset$Province == "PrinceEdwardIsland", "Location"] = "Maritime"
subset[subset$Province == "Quebec", "Location"] = "East"
subset[subset$Province == "Saskatchewan", "Location"] = "Prairie"


table(subset$Province, subset$Location)







```


## Time Series Clustering
```{r, echo=FALSE}
install.packages("TSclust")
library(TSclust)
data("synthetic.tseries")
head(synthetic.tseries)
# LONG TO WIDE format
wide_data = reshape(subset, idvar = c("When"), timevar="Province", direction="wide")
View(wide_data)
ts = wide_data[colnames(wide_data)[agrep('Unemp',colnames(wide_data))]]
dist = diss(ts, "INT.PER")
clusters = cutree(hclust(dist), k=5)
c_names = names(clusters)
ts_t = data.frame(t(ts))

for (c in c_names){
  print(c)
  print(clusters[c])
  ts_t[c,]$cluster = clusters[[c]]
  ts_t[c,'province'] = c
  print(ts_t[c,])
}

# plot together with cluster result
ts = melt(ts_t, id.vars=c('province', 'cluster'))



interaction.plot()
```

idx = seq(0, 6.28, len=100)
query = sin(idx)+runif(100)/10
temp = cos(idx)
idx
query
